Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar

### 2018-12-03 accessPEP.Rmd
Changing to simpler model and cleaning data using dplyr

---
title: "Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar"
author: ""
date: 
output:
  word_document:
    reference_docx: format_docs/word_styles_reference_01.docx
  pdf_document:
    includes:
      in_header: mystyle.sty
  html_document: default
css: style.css
---

```{r directory, echo=FALSE, message = FALSE, results = 'hide'}
rm(list = ls())
require("knitr")
knitr::opts_chunk$set(fig.pos = 'h')
```

@@

Human rabies post-exposure prophylaxis prevents fatalities if delivered in a timely manner, however, access to this intervention is highly limited in areas where the disease is endemic. Canine rabies causes an estimated 59,000 human deaths annually that could be prevented by PEP.

@@

The majority of studies which estimate burden of rabies use probability decision tree frameworks, often with the key assumptions that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) are proportional to rabies incidence and that reporting is uniform across space. However, these estimations… Given little data on true exposures of rabies how do we do this? Newly developed frameworks which rely on in-country data on key parameters likely better capture burden by disentangling rabid from non-rabid bites and generating incidence from estimates of the dog population and control in the dog population.

Here we extend this framework to include spatial access to PEP, using bite patient data from 27 clinics in Madagascar, a rabies endemic country with a largely absent control program in the canine population. Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to the government or to the patients at 31 clinics across the country. PEP is not available at any other public clinics or through the private sector. Given that rabies is endemic in the country and there is effectively no vaccination of domestic dogs, we use predictions of incidence from dynamic models and 
We flip the standard decision tree and make the assumption that reported bite incidence reflects access and reporting to PEP rather than differences in rabies incidence.

@@

### Bite patient data
We used the database of bite patient forms submitted from 27 ARMC between 2011-2014. These were paper forms that were sent to IPM as frequently as monthly to annually by individual clinics. Two clinics, the IPM ARMC and the Fianarantsoa ARMC used computer databases from which the data during this period were extracted and merged to the larger database. These data include details on the location of the bite patient and biting animal (*cite baseline data paper*). We also had the number of doses delivered on an annual basis to each clinic and used this information to estimate submission rates of paper forms to IPM for each clinic as described in (*Moramanga paper + baseline data paper*, using minimum # of doses expected based on clinic throughput). As there were many discrepancies between commune names in the GIS administrative layer and the names written on the forms, we used fuzzy matching to find the closest matching commune names and manually checked these.

@@

### GIS Data
We use the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al. 2015,) and GPS points of clinics to get travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale. We then averaged these to the commune level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs. For each clinic, we defined the catchment area as all communes for which the clinic was the closest ARMC. Population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al. 2012) and also aggreagated to the commune level.

@@

### Model of reporting as a function of travel time
We model the bites reported annually to the clinic ($B_{i,t}$) for each commune using a Bayesian state-space framework as follows:

$$ B_{i,t} =  k_{c,t} \times \rho_i \times(R_{i,t} + N_{i,t}) $$

$k_{c,t}$ is the annual reporting at the catchment level (the product of the estimated proportion of forms that were submitted and the proportion of patients for which we were able to match commune names for that year).

$p_i$ is the reporting of all bites which follows the logistic function:

$$ \rho_{i} = \frac{C}{1 + e^{\frac{A - x}{d}}}$$

where $C$ is a limiting function on the y-axis that measures probability of attendance when distance is zero, $X$ is the log transformed travel times in minutes, $d$ is the distance decay parameter, and A is an asymptote factor at an inflection point of the model. We fix C to 0.90 (estimated from the Moramanga data), as even at minimal travel times, there are people that do not report bites for other reasons than accessibility.

@@

$N_{i,t}$ is the non-rabid bite incidence as a function of human population size **(Need to think about how to justify this...)**.

$$ N_{i,t} = Poisson(\beta_pP_i)$$

$R_{i,t}$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures. We calculate this as:

@@

where $r$ is rabies incidence in dogs in the absence of any vaccination multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al. 2018). We estimate the dog population by using a human:dog ratio of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence. As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range of HDRs observed across Africa(cite?).

@@

We use RJAGS (...) to fit this model to our reported bites at the commune level over the 4-year period using non-informative priors for $A$ $d$ and $\beta_p$

@@

### Model evaluation and prediction
We fitted the model to data from n catchments which had above x% clinic level reporting {$k_c$} for all four years.

We used

For three other catchments with reporting above x% we used to out-of-fit predict...
Ability to predict average annual/monthly bite incidence by travel time
Ability to predict out-of-fit to other catchments 
(observed to expected mean incidence over 4 year period (monthly or annual)

### Sensitivity analyses
Sensitivity to R (@ what level of human exposure incidence!)
Simulate and recover params? @ what *true* incidence does p_i become non-identifiable...

### Exploring alternative scenarios and estimation of burden
We estimate burden of human rabies deaths as follows:

$$ D_i = p_{death}\times (1-\rho_i) \times R_i $$
where $p_{death}$ is the probability of death given an exposure in the absence of PEP, $\rho_i$ is the estimated commune specific reporting probability derived from the model, and $R_i$ is the expected rabies incidence. **We use Monte Carlo? simulation drawing from a dist of incidence to get an average number of expected deaths.** We do not include potential deaths due to imperfect PEP (i.e. not completing at least 3 doses), delays in adminsitration, or lack of RIG.

We consider three scenarios:
1. The baseline with the current clinic locations (n = 31)
2. Expansion of ARMC to one clinic per district (n = 114)
3. Expansion of ARMC to all CSB IIs (n = ...?)

We use data on the location of **CSBs** provided by IPM to regenerate travel times to the nearest ARMC given expansion as per scenario 2 and 3. We then predict the expected reporting rate from the model given these new travel times and compare the relative decreases in burden for the three scenarios.

>>>>

---
title: "Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar"
author: ""
date: 
output:
  html_document: default
  pdf_document:
    includes:
      in_header: mystyle.sty
  word_document:
    reference_docx: format_docs/word_styles_reference_01.docx
css: style.css
---

```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
rm(list = ls())
library(knitr)
library(RImagePalette)
library(jpeg)
library(tidyverse)
library(lubridate)
library(magrittr)
library(rgdal)
library(broom)
knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, results = 'hide')

```

@@

Canine rabies causes an estimated 59,000 human deaths annually. Post-exposure prophylaxis (PEP) prevents fatalities if delivered in a timely manner, however, access to this intervention is highly limited in areas where the disease is endemic. Data on true rabies exposures in humans and incidence in animals is also lacking in most countries. Often, the only data available are numbers of bite victims reporting to health facilities. The majority of rabies burden studies use these data to estimate burden of rabies from probability decision tree frameworks, often with the key assumptions that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) are proportional to rabies incidence and that reporting is uniform across space.

The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on level of vaccination coverage and size of the dog population at the national level. Using transmission dynamic models to estimate incidence could improve upon previous studies which may overestimate rabies burden in areas of high reporting and underestimate in areas with low reporting due to the assumption that overall reported bite incidence is proportional to rabies incidence.

@@

Here, we extend this framework to include spatial access to PEP, using bite patient data from 27 clinics in Madagascar, a rabies endemic country that lacks a national strategy for controlling rabies in domestic dogs. Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to the government or to patients that present to 31 clinics across the country. PEP is not available at any other public clinics or through the private sector. Given that rabies is endemic in the country and there is effectively no vaccination of domestic dogs, we use the baseline predictions of incidence from the dynamic model reported in Hampson et al. 2018. We flip the standard decision tree and make the assumption that reported bite incidence reflects access and reporting to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of bites reported to clinics.

@@

### Bite patient data
We used the database of bite patient forms submitted from 24 ARMC between 2014 - 2017. These were individual patient data forms that were submitted to IPM as frequently as monthly to annually by clinics. Two clinics, the IPM ARMC and the Fianarantsoa ARMC used computer databases from which the data during this period were extracted and merged to the larger database. These data include the administrative district of the bite patient's address. As previous analyses showed likely substantial undersubmission of forms from clinics (*cite baseline data paper*), we calculated the proportion of days of the year for which forms were submitted. We excluded periods with 10 or more consecutive dates with no patient forms and calculated the proportion of days included per year. Figure 1a + 1b?

@@

### GIS Data
We use the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al. 2015,) and GPS points of clinics to get travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale. We then averaged these to the commune level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs. For each clinic, we defined the catchment area as all communes for which the clinic was the closest ARMC. Population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al. 2012) and also aggreagated to the commune level. Figure 2 = networky fig + catchments!

@@

### Model of reported bites as a function of travel time
Excluding any catchments for which there was less than 25% estimated form submission, we calculated the average bite incidence corrected by the annual undersubmission of forms at the district level.

We fit a linear mixed effects model where the annual average bite incidence at the district level was predicted by the average travel time to the nearest CTAR for the district, the population within the district, and a random catchment level effect.

For three other catchments with reporting above x% we used to out-of-fit predict...
Ability to predict average annual/monthly bite incidence by travel time
Ability to predict out-of-fit to other catchments

@@

### Estimation of burden and reporting 
We used our model to predict average annual bite incidence for all 114 districts in Madagascar, and estimated deaths due to rabies given this and assumptions about rabies exposures.

We calculated deaths as:
$$deaths_{i} = (R_{i} - ({p_{rabid} \times B_{i})) \times p_{death}$$

where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
$$
           p_{rabid}=
            \begin{cases}
            x, & \text{if}\ \frac{R_i \times \rho_{max}}{B_i} > x \\
            \frac{R_i}{B_i}, & \text{otherwise}
            \end{cases}
$$
<br>

such that rabid reported bites (i.e. $p_{rabid}\timesB_{i}$) cannot exceed the expected number of human exposures given maximum reporting (i.e. $R_{i} \times \rho_{max}$. $p_{rabid}$ is taken from Rajeev et al. 2018 on contact tracing data from the Moramanga District.

$R_{i}$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures. We calculate this as:

@@

where $r$ is rabies incidence in dogs in the absence of any vaccination multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al. 2018). We estimate the dog population by using a human:dog ratio of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence. As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range observed across Africa(cite?).

@@

## Estimating the impact of expanding PEP Access
We use this framework to compare three scenarios of PEP provisioning in Madagascar:
1. The baseline with the current clinic locations (n = 31)
2. Expansion of ARMC to one clinic per district (n = 114)
3. Expansion of ARMC to all CSB IIs (n = ...?)

We use data on the location of **CSBs** provided by IPM to regenerate travel times to the nearest ARMC given expansion as per scenario 2 and 3. We then predict the expected bite incidence from the model given these new travel times and compare the relative decreases in burden for the three scenarios.

@@

### Sensitivity Analyses
- Sensitivity to assumptions of R
- Sensitivity to assumptions of p_rabid

## Preliminary data processing

### Cleaning data
```{r first cleaning}
## read in data
ctar_data <- read.csv("data/SaisieRage_DATA_2018-09-21_1755.csv")

## percent with wrong dates by ctar

## percent with missing district by ctar

## exclude contacts

```

## Submission of forms
- Missingness seems to have been due to periods of time without form submission
- Calculating periods of time

```{r doses}
ctar_metadata <- read.csv("data/ctar_metadata.csv")
mada_district <- readOGR("data/MadaGIS/MadaPops.shp")

img <- readJPEG("~/Downloads/Dp0j4GNUwAAPPJV.jpeg")
pal <- image_palette(img, 31)
ctar_metadata$color <- pal

ctar_metadata %>% 
  dplyr::select(id_ctar, starts_with("doses")) %>%
  gather(year, doses, -id_ctar) %>%
  mutate(year = as.numeric(substr(as.character(year), 7, 10))) -> doses

## getting dose time series by clinic
ctar_data %>%
  select(id_ctar, datej0, datej3, datej7, datej28) %>%
  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>%
  gather(dose, date, -id_ctar) %>%
  group_by(id_ctar, date) %>%
  summarise(n = n()) -> dose_ts
dose_ts <- dose_ts[!is.na(dose_ts$date), ]
dose_ts$ctar <- ctar_metadata$CTAR[match(dose_ts$id_ctar, ctar_metadata$id_ctar)]
dose_ts <- dose_ts[!is.na(dose_ts$ctar), ]

## getting theoretical # vials by year
dose_ts %>%
  mutate(year = year(date), vials = ceiling(n/2)) %>%
  group_by(id_ctar, year) %>%
  summarize(vials = sum(vials)) %>%
  left_join(doses) -> dose_comp

ggplot(data = dose_ts, aes(date, ctar)) + 
  geom_tile(aes(fill = n)) +
  scale_fill_continuous(type = "viridis", name = "Patients")

## getting reporting
start_date <- ymd("2014-01-01")
end_date <- ymd("2017-12-31")
ts <- as_tibble(seq(start_date, end_date, by = "day"))

dose_ts %>%
  right_join(ts, by = c("date" = "value")) %>%
  select(id_ctar, date, n) %>%
  spread(id_ctar, n) %>%
  replace(., is.na(.), 0) -> doses_wide
dose_mat <- as.matrix(doses_wide[, 2:(ncol(doses_wide) - 1)])

date_mat <- matrix(NA, nrow(dose_mat), ncol(dose_mat))
for (j in 1:ncol(date_mat)){
  rle(dose_mat[ , j]) %>%
    unclass() %>%
    as.data.frame() %>%
    mutate(end = cumsum(lengths),
         start = c(1, lag(end)[-1] + 1)) %>%
    filter(values == 0, lengths >= 10) -> rles
  for (i in 1:nrow(rles)){
    date_mat[rles$start[i]:rles$end[i], j] <- 0
  }
}

date_mat <- replace(date_mat, is.na(date_mat), 1)
date_mat %>% 
  as_tibble() %>%
  group_by(date = year(doses_wide$date)) %>%
  summarise_all(funs(sum(.)/365)) -> clinic_reporting
colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]
clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date)
clinic_reporting$ctar <- ctar_metadata$CTAR[match(clinic_reporting$id_ctar, 
                                                  ctar_metadata$id_ctar)]

ggplot(clinic_reporting, aes(x = date, y = ctar, fill = prop)) + 
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue")

to_exclude <- c(grep("contact", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("contam", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("consom", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("passage", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE))
to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
ctar_data$excl <- 0
ctar_data$excl[ctar_data$remarque %in% to_exclude] <- 1

ctar_data %>% 
  filter(excl != 1) %>% # exclude contacts
  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
  group_by(year = year(date_de_consultation), district, id_ctar) %>% # group by year and district
  summarize(n = n()) -> exposures # get count of exposures

## districts that are non-reporter and data not available yet CTAR
no_data <- c("Manja", 'Ambatondrazaka', "Vangaindrano", "Tsiroanomandidy",
             "Antsiranana I", "Taolagnaro", "Ambovombe-Androy", "Nosy-Be", "Ambatomainty",
             "Marolambo", "Sainte Marie", "Antananarivo Renivohitra", "Antananarivo Avaradrano",
             "Ambohidratrimo", "Ankazobe", "Manjakandriana", "Anjozorobe", "Andramasina", 
             "Antananarivo Atsimondrano", "Fianarantsoa I", "Ambalavao", "Ambohimahasoa", 
             "Ikalamavony", "Lalangina", "Vohibato", "Isandra")
no_data <- mada_district$mdg_dis_co[match(no_data, mada_district$district)]
mada_district@data %>%
  select(district = mdg_dis_co, pop = pop2015adj, ttimes = study_area) -> covars

exposures %>%
  rename(date = year) %>%
  mutate(id_ctar = as.character(id_ctar)) %>%
  left_join(clinic_reporting) %>%   # add reporting by clinic + year
  filter(!(district %in% no_data), prop >= 0.25) %>% # exclude no data & < 25% annual reporting
  group_by(district) %>%
  mutate(bites = n/prop) %>%
  summarise(bites = sum(bites)) %>% # average by ctar + district 
  left_join(covars) %>% # add pop + ttimes + reporting
  mutate(incidence = bites/(pop/1e5)) -> exps_cleaned 
exps_cleaned %$% plot(ttimes, bites)
exps_cleaned %$% plot(ttimes, incidence, bty = 
                        "n", pch = 20)

mada_dist_gg <- tidy(mada_district, region = "mdg_dis_co")
mada_dist_gg <- left_join(mada_dist_gg, exps_cleaned, by = c("id" = "district"))

ggplot() +
  geom_polygon(data = mada_dist_gg, aes(x = long, y = lat, group = group, fill = incidence)) +
  scale_fill_continuous(type = "viridis") +
  geom_point(data = ctar_metadata, aes (x = LONGITUDE, y = LATITUDE))
ggplot() +
  geom_polygon(data = mada_dist_gg, aes(x = long, y = lat, group = group, fill = ttimes)) +
  scale_fill_continuous(type = "viridis") +
  geom_point(data = ctar_metadata, aes (x = LONGITUDE, y = LATITUDE))

library(lme4)
summary(lm(log(incidence) ~ ttimes, data = exps_cleaned))
```
